Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-10-13 manuscript2.RMD
proofreading

<div custom-style="correspondence_style">
^1^Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway
</div>

<div custom-style="correspondence_style">
Correspondance: Kristian Lian (kristian.lian@inn.no), Stian Ellefsen (stian.ellefsen@inn.no)
</div>

ORCID: Kristian Lian https://orcid.org/0000-0002-5169-5960, Daniel Hammarström https://orcid.org/0000-0001-8360-2100, Håvard Hamarsland https://orcid.org/0000-0002-7343-2774, Stian Ellefsen https://orcid.org/0000-0003-0467-2018
</div>

<div custom-style="abstract_style">
## Abstract

<!-- To do: Fix tables, add outlier figures to the readme -->
```{r}
#| echo: false

@@

**Conclusion:** Glucose ingestion before and after resistance training did not augment ribosomal RNA accumulation during two weeks of heavy-load resistance training in moderately trained young adults.

@@

### Acknowledgements
Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak and Thea Bøhn Nyløkken for their dedication and hard work during the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

**Abbreviations**
```{r abbreviation table}
### Abbreviations table

@@

abb.tab
```

\pagebreak
## Introduction

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. There may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], however, the internal physiological milieu seems favourably shaped for muscle growth by factors such as different types of training stress and nutrients [@figueiredo_genetic_2021; @sebata_control_2018; @thalacker-mercer_cluster_2013]. Indeed, training stress and nutrients impact the ability to synthesise ribosomes, which in turn is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020; @kusnadi_regulation_2015; @sebata_control_2018]. For instance, increasing training volume generally induces greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017]. Still, as evident from Hammarström et al. (2020), not all participants experience increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine are often used to optimise RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains equivocal if other nutritional adjuvants such as glucose can increase the efficacy of RT [@tezze_anabolic_2023]. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. In addition, insulin may have anabolic effects with elevated amino acids, and has a clear role in the reduction of muscle protein breakdown independent of amino acid availability [@abdulla_role_2016; @hillier_physiological_2000]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Ribosomal biogenesis and content seem to be a prerequisite for skeletal muscle growth, and transcription of ribosomal RNA (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995]. Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target ribosomal protein S6 kinase beta-1 (S6K1) serves as a direct mediator of insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019; @von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Indeed, UBF phosphorylation, which is required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, seems to be increased by high-glucose treatment in a mTORC1-dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011]. Independent of UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures [@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.
<!-- sørg for at disse to avsnittene henger godt sammen!-->
There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency concerning long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucoseum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and -c-peptide levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours (hrs) after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose *or* protein and placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs. 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of protein supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. During sessions, participants were free to ingest water ad libitum.

@@

Strength tests were performed before (Figure 1A, Days -7 and -5, and -1, both legs) and during the intervention (Figure 1A, on days 4 and 8 for leg 1, and days 5 and 9 for leg two), after session 5 and after finalization of the intervention (Figure 1A, on days 11/12 for leg 1 and days 12/13 for leg 2). Maximal isometric and isokinetic knee extension torque was measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massachusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (Figure 1A, days -7 and -5). Isokinetic peak torque was measured concentrically from 90 to 0 degrees knee angle (extended knee was set to 0 degrees) at angular velocities of 60- and 240 degrees per second, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric knee extensor peak torque was measured at a kneen angle of 60 degrees, for a maximum of 10 seconds and two repetitions per test. The isometric tests were ended when the participants reached a plateau or peak torque development decreased, which on average occurred between 2-4 seconds into the test. During days 4, 5, 8 and 9 (Figure 1D, days 4 and 5 = Post 2RT, days 8 and 9 = Post 4RT), humac tests were conducted one hour before RT on the leg performing RT the previous day. During days 11 and 12, humac tests were performed four times: I) 45min before the last RT session (Figure 1D, Post 5RT), II) 30min after the last RT session (Figure 1D, 30 min post 6RT), III) 2hrs after the last RT session (Figure 1D, 2h post 6RT), and IV) 23hrs after the last RT session (Figure 1D, 23h post 6RT). Test I on day 12/13 included testing of both legs, representing 23hrs post-RT session test of one leg and post-session 5 test of the other leg. The highest peak torque values from the respective angular velocities and time points were summarized in an index. The index was calculated by dividing the average peak torque value by the highest observed peak torque value per angular velocity and summarizing this new index per angular velocity to a mean strength index.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at before the intervention during familiarization (Figure 1, Days -7 and -5). The participants performed a general warm-up with 10 minutes of cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to ~50-75% of assumed max repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, starting from a knee angle of 90 degrees. To find a reproducible 90-degree knee angle for each participant, centimetre markings on the side panels of the leg press machine were used to record where to find 90 degrees for each separate leg and participant. Attempts where participants did not reach 90 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up as the maximal leg press test and was defined as the maximal load lifted in a controlled fashion, reaching full extension of the knee joint.  Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of unilateral knee extensions, with an intensity of 10 repetitions maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10 minutes. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at ~50% and ~70% of 10RM. To ensure adequate exercise stimulation throughout the intervention, the exercise load was increased the following set if the participants lifted more than 12 repetitions, as a progressive loading strategy. If the participants lifted fewer than 8 repetitions per set, the load was reduced in the following set. The resting time between working sets was two minutes. For safety and standardisation purposes, all sessions were monitored by trained personnel. Lastly, training volume (load and repetitions) was logged for every session.

@@

Muscle biopsies were sampled from m. vastus lateralis using well-established procedures [@hammarstrom_benefits_2020]. Briefly, muscle biopsy sampling was performed under local anaesthesia (Xylocaine, 10 mg ml^-1^ with adrenaline 5 μg ml^−1^, AstraZeneca AS, Oslo, Norway) using a 12-gauge needle (Universal Plus, Mermaid Medical AS, Stenløse, Denmark), operated with a spring-loaded biopsy gun (Bard Magnum, Bard, Rud, Norway). After the biopsy sampling, muscle tissue was divided into two aliquots for determination of total RNA/expression of rRNA and two aliquots for protein content measurement. Aliquots were snap-frozen in isopentane (-80°C) and stored at -80°C until further analyses. Muscle biopsies were collected at four time points: I/II) Before the intervention (Figure 1A, 2hrs before training, Day 1 = leg 1, Day 2 = leg 2), and III/IV) after the fifth RT session, two hours before the sixth RT session (Figure 1A, Day 11 = leg 1, Day 12 = leg 2). At each time point, two samples were taken from the same incision. To standardize this procedure, all individual participants had biopsies taken at the same time of day, in an overnight fasted state.

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger draws on days with biopsy sampling. One capillary blood sample was collected on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately before protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5 hours after protein ingestion (0830hrs, i.e., immediately before GLU/PLAC intake), IV) 2 hours after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (~0915hrs), V) immediately after training (~0930hrs), and VI) 2 hours after completion of training (~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45 minutes after protein ingestion and in the middle of the RT session, to analyse endocrine variables.

### Total RNA extraction and real-time reverse transcription polymerase chain reaction
```{r totrna-extraction}

@@

Two muscle biopsy aliquots were used for total RNA extraction per leg, resulting in a total of eight RNA samples per participant. Total RNA was extracted using TRIzol with muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). To enable analysis of target gene expression per unit tissue weight [@ellefsen_improved_2008; @ellefsen_reliable_2014], an exogenous RNA control (Lambda, λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount to each sample (0.04 ng ml^-1^ of TRIzol reagent). For assessment of RNA content and purity, RNA was eluted in TE buffer (1:2) and assessed via spectrophotometry. All samples had a 260 nm to 280 nm ratio > 1.9. The RNA stock was stored at -80°C until further analyses. Before quantitative analyses of total RNA, samples with known loss of RNA during extraction (n = `r technical`) or a deviation from the observed RNA to muscle tissue weight relationship larger than 3 $\times$ residual SD while accounting for training status (n = `r observed`) were removed from the data set. Total RNA was normalised to wet muscle weight and log transformed before statistical analyses.

@@

Glucose ingestion compared to placebo before and after RT sessions did not lead to a difference in mean change of total training session volume, with an average increase of 18 and 16% from session 1 to session 6, respectively (*p* > 0.05). There were no differences in mean macronutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* > 0.05 for all).

@@

Complete data sets and scripts are downloadable here; https://github.com/Kristianlian/ribose-paper

@@

Extended methods can be found here; https://github.com/Kristianlian/ribose-paper

\pagebreak
## References

<div id="refs"></div>

@@

**Table 1. Participant characteristics:** Values are means ± SD. 60º sec-1 = 60 degrees per second movement speed (isokinetic), 240º sec-1 = 240 degrees per second movement speed (isokinetic), 0º sec-1 = 0 degrees per second movement speed (isometric).
```{r baseline table, echo = FALSE}

>>>>

::: {custom-style="correspondence_style"}
^1^Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway
:::

::: {custom-style="correspondence_style"}
Correspondance: Kristian Lian ([kristian.lian\@inn.no](mailto:kristian.lian@inn.no){.email}), Stian Ellefsen ([stian.ellefsen\@inn.no](mailto:stian.ellefsen@inn.no){.email})
:::

ORCID: Kristian Lian <https://orcid.org/0000-0002-5169-5960>, Daniel Hammarström <https://orcid.org/0000-0001-8360-2100>, Håvard Hamarsland <https://orcid.org/0000-0002-7343-2774>, Stian Ellefsen <https://orcid.org/0000-0003-0467-2018>

::: {custom-style="abstract_style"}
## Abstract

<!-- To do: Fix tables, add outlier figures to the readme -->

```{r}
#| echo: false

@@

**Conclusion:** Glucose ingestion before and after resistance training did not augment ribosomal RNA accumulation during twelve days of heavy-load resistance training in moderately trained young adults.

@@

### Acknowledgements

Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak and Thea Bøhn Nyløkken for their dedication and hard work during the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

**Abbreviations**

```{r abbreviation table}
### Abbreviations table

@@

abb.tab
```
:::

\pagebreak

## Introduction

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. There may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], however, the internal physiological milieu seems favourably shaped for muscle growth by factors such as different types of training stress and nutrients [@figueiredo_genetic_2021; @sebata_control_2018; @thalacker-mercer_cluster_2013]. Indeed, training stress and nutrients impact the ability to synthesise ribosomes, which in turn is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020; @kusnadi_regulation_2015; @sebata_control_2018]. For instance, increasing training volume generally induces greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020; @krieger_single_2009; @schoenfeld_strength_2017]. Still, as evident from Hammarström et al. (2020), not all participants experience increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine are often used to optimise RT adaptations [@cermak_protein_2012; @lanhers_creatine_2015; @lanhers_creatine_2017; @morton_systematic_2018]. However, it remains equivocal if other nutritional adjuvants such as glucose can increase the efficacy of RT [@tezze_anabolic_2023]. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015; @sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015; @moss_housekeeper_2007; @sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019; @hammarstrom_benefits_2020; @sebata_control_2018; @stec_ribosome_2016; @von_walden_ribosome_2019]. In addition, insulin may have anabolic effects with elevated amino acids, and has a clear role in the reduction of muscle protein breakdown independent of amino acid availability [@abdulla_role_2016; @hillier_physiological_2000]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Ribosomal biogenesis and content seem to be a prerequisite for skeletal muscle growth, and transcription of ribosomal ribonucleic acid (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995]. Multiple signalling pathways converge to regulate rRNA transcription, including c-Myc and the mammalian target of rapamycin complex 1 (mTORC1) signal-transduction pathway [@kusnadi_regulation_2015; @von_walden_ribosome_2019; @sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. The general transcription factor c-Myc increases ribosomal biogenesis directly through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011; @mori_c-myc_2021; @west_acute_2016]. Indeed, UBF phosphorylation, which is required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, seems to be increased by high-glucose treatment in a mTORC1-dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011]. Independent of UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures [@zhai_human_2012]. The mTORC1 pathway receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009] and contributes to ribosome biogenesis through the forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription and through the regulation of ribosomal protein translation [@figueiredo_regulation_2019; @von_walden_ribosome_2019]. Moreover, mTORC1 and its downstream target ribosomal protein S6 kinase beta-1 (S6K1) serves as direct mediators of insulin signalling in skeletal muscle [@hillier_physiological_2000]. Together these observations indicate a potential role of glucose in positively affecting human skeletal muscle rRNA synthesis, through potentiating transcription or translation of ribosomal RNA and increasing the translational capacity of skeletal muscle cells [@sebata_control_2018; @zhai_human_2012; @hillier_physiological_2000; @hoppe_amp-activated_2009].

There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency concerning long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011; @von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

@@

The study was designed as a 12-day double-blinded placebo-controlled simultaneous crossover trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucoseum monohydricum vs. \~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and serum c-peptide levels). Participants were asked to avoid resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours (hrs) after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose *or* protein and placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs. 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (\~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (\~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of protein supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. During sessions, participants were free to ingest water ad libitum.

@@

Strength tests were performed before (Figure 1A, Days -7 and -5, and -1, both legs) and during the intervention (Figure 1A, on days 4 and 8 for leg 1, and days 5 and 9 for leg two), after session 5 and after finalization of the intervention (Figure 1A, on days 11/12 for leg 1 and days 12/13 for leg 2). Maximal isometric and isokinetic knee extension torque was measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massachusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (Figure 1A, days -7 and -5). Isokinetic peak torque was measured concentrically from 90 to 0 degrees knee angle (extended knee was set to 0 degrees) at angular velocities of 60- and 240 degrees per second, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric knee extensor peak torque was measured at a knee angle of 60 degrees, for a maximum of 10 seconds and two repetitions per test. The isometric tests were ended when the participants reached a plateau or peak torque development decreased, which on average occurred between 2-4 seconds into the test. During days 4, 5, 8 and 9 (Figure 1D, days 4 and 5 = Post 2RT, days 8 and 9 = Post 4RT), humac tests were conducted one hour before RT on the leg performing RT the previous day. During days 11 and 12, humac tests were performed four times: I) 45min before the last RT session (Figure 1D, Post 5RT), II) 30min after the last RT session (Figure 1D, 30 min post 6RT), III) 2hrs after the last RT session (Figure 1D, 2h post 6RT), and IV) 23hrs after the last RT session (Figure 1D, 23h post 6RT). Test I on day 12/13 included testing of both legs, representing 23hrs post-RT session test of one leg and post-session 5 test of the other leg. The highest peak torque values from the respective angular velocities and time points were summarized in an index. The index was calculated by dividing the average peak torque value by the highest observed peak torque value per angular velocity and summarizing this new index per angular velocity to a mean strength index.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at before the intervention during familiarization (Figure 1, Days -7 and -5). The participants performed a general warm-up with 10 minutes of cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to \~50-75% of assumed max repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, starting from a knee angle of 90 degrees. To find a reproducible 90-degree knee angle for each participant, centimetre markings on the side panels of the leg press machine were used to record where to find 90 degrees for each separate leg and participant. Attempts where participants did not reach 90 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up as the maximal leg press test and was defined as the maximal load lifted in a controlled fashion, reaching full extension of the knee joint. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of unilateral knee extensions, with an intensity of 10 repetitions maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10 minutes. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at \~50% and \~70% of 10RM. To ensure adequate exercise stimulation throughout the intervention, the exercise load was increased the following set if the participants lifted more than 12 repetitions, as a progressive loading strategy. If the participants lifted fewer than 8 repetitions per set, the load was reduced in the following set. The resting time between working sets was two minutes. For safety and standardisation purposes, all sessions were monitored by trained personnel. Lastly, training volume (load and repetitions) was logged for every session.

@@

Muscle biopsies were sampled from m. vastus lateralis using well-established procedures [@hammarstrom_benefits_2020]. Briefly, muscle biopsy sampling was performed under local anaesthesia (Xylocaine, 10 mg ml^-1^ with adrenaline 5 μg ml^−1^, AstraZeneca AS, Oslo, Norway) using a 12-gauge needle (Universal Plus, Mermaid Medical AS, Stenløse, Denmark), operated with a spring-loaded biopsy gun (Bard Magnum, Bard, Rud, Norway). After the biopsy sampling, muscle tissue was divided into two aliquots for determination of total RNA/expression of rRNA and two aliquots for protein content measurement. Aliquots were snap-frozen in isopentane (-80°C) and stored at -80°C until further analyses. Muscle biopsies were collected at four time points: I/II) Before the intervention (Figure 1A, 2hrs before training, Day 1 = leg 1, Day 2 = leg 2), and III/IV) approximately 22 hours after the fifth RT session, two hours before the sixth RT session (Figure 1A, Day 11 = leg 1, Day 12 = leg 2). At each time point, two samples were taken from the same incision. To standardize this procedure, all individual participants had biopsies taken at the same time of day, in an overnight fasted state.

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger draws on days with biopsy sampling. One capillary blood sample was collected on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately before protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5 hours after protein ingestion (0830hrs, i.e., immediately before GLU/PLAC intake), IV) 2 hours after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (\~0915hrs), V) immediately after training (\~0930hrs), and VI) 2 hours after completion of training (\~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45 minutes after protein ingestion and in the middle of the RT session, to analyse endocrine variables.

### Total RNA extraction and real-time reverse transcription polymerase chain reaction

```{r totrna-extraction}

@@

Two muscle biopsy aliquots were used for total RNA extraction per leg, resulting in a total of eight RNA samples per participant. Total RNA was extracted using TRIzol with muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (\~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). To enable analysis of target gene expression per unit tissue weight [@ellefsen_improved_2008; @ellefsen_reliable_2014], an exogenous RNA control (Lambda, λ polyA External Standard Kit, Takara Bio Inc., Shiga, Japan) was added at a fixed amount to each sample (0.04 ng ml^-1^ of TRIzol reagent). For assessment of RNA content and purity, RNA was eluted in TE buffer (1:2) and assessed via spectrophotometry. All samples had a 260 nm to 280 nm ratio \> 1.9. The RNA stock was stored at -80°C until further analyses. Before quantitative analyses of total RNA, samples with known loss of RNA during extraction (n = `r technical`) or a deviation from the observed RNA to muscle tissue weight relationship larger than 3 $\times$ residual SD while accounting for training status (n = `r observed`) were removed from the data set. Total RNA was normalised to wet muscle weight and log transformed before statistical analyses.

@@

Glucose ingestion before and after RT sessions did not lead to a higher mean change of total training session volume, with a mean increase of 17% in both the glucose condition (pre: 5262 ± 1799kg, post: 6319 ± 2256kg, *p* > 0.05) and the placebo condition (pre: 5351 ± 1615kg, post: 6438 ± 2092, *p* > 0.05) from session 1 to session 6, respectively. There were no differences in mean macronutrient intake (protein, fat, carbohydrates) or total calorie intake between glucose and placebo on pairwise consecutive days (*p* \> 0.05 for all, raw data and table available at GitHub repository).

@@

Complete data sets and scripts are downloadable here; <https://github.com/Kristianlian/ribose-paper>

@@

Extended methods can be found here; <https://github.com/Kristianlian/ribose-paper>

\pagebreak

## References

::: {#refs}
:::

@@

**Table 1. Participant characteristics:** Values are means ± SD. 60º sec-1 = 60 degrees per second movement speed (isokinetic), 240º sec-1 = 240 degrees per second movement speed (isokinetic), 0º sec-1 = 0 degrees per second movement speed (isometric).

```{r baseline table, echo = FALSE}
